Baseline centering problem

### 2018-12-13 paper/tangotd.tex

\newcommand{\Dir}{\mathcal{D}}
\newcommand{\green}{\mathcal{G}}

@@

$G$ cannot be stored but one can store an approximate model. For instance
one could approximate $G$ via a neural network with input $s$ and output
in the policy parameter space. \footnote{The policy parameter space may
be quite large. This means the neural network should have a small
next-to-last layer. Indeed, with a next-to-last layer of size $r$,
the weights of the last layer are represented by $r$ policy parameters.
This corresponds to a rank-$r$ approximation of $G$,
as a sum of parameters times functions:
\begin{equation}
G\approx \sum_i \psi_i \phi_i(s)
\end{equation}
where each $\psi_i$ is a policy parameter, and each $\phi_i$ is
a function of the state, corresponding to a feature in the network.} Then the reverse
Bellman equation above should allows us to learn this model via
a TD-like algorithm, \emph{on-policy}. TODO: HOW? not clear how to
sample from $\phi P$ if $\phi$ is a function. Would need a model of
reverse transitions? \footnote{I think one has to learn $G\mu^{-1}$ and
the equation $G=\Id+GP$ rewrites as $\mu
(G\mu^{-1})\mu=\mu+\mu(G\mu^{-1})\mu P$ where everything is samplable.
However I fear an implicit double sampling in there; probably
featurize/dualize
by asking that this is satisfied for any functions on the left and right.
}

@@

\paragraph{A learned representation for $\Delta^{-1}$.} $\Delta^{-1}$
contains all the necessary to do policy gradient or actor-critic with
arbitrary reward functions, as seen above. Is it feasible to store a
compact and useful representation of $\Delta^{-1}$? Ideally, such a
representation should be learned from the \emph{two-sided} Bellman
equations
\begin{equation}
\label{eq:greenbell}
\Delta^{-1}=\Id + P \Delta^{-1},\qquad \Delta^{-1}=\Id + \Delta^{-1}P
\end{equation}
With these two equations, when observing a transition $s\to s'$, we update
both the model of what happens when we start at $s$ (left equation,
standard Bellman), and the model of how to arrive at $s'$ (right
equation, rarely used as far as I know).

The left Bellman equation on $\Delta^{-1}$ is equivalent to
simultaneously solving the Bellman equation with all possible reward
functions, generated by $\1_{s'}$ for all possible target states $s'$.

``Universal value function approximators'' (Schaul--Horgan--Gregor--Silver ICML 2015) do a bit of that, but only
with the left (ordinary) Bellman equation. It amounts to solving a number
of
ordinary Bellman equations with different reward functions, and relying on the generalization
abilities of the network, but the right equation is not exploited at all.

We propose to learn two functions $\green(s)$ and $f(s)$ from states into
a state representation $E$, such that for any test function $\psi\from
S\to \R$, for any $s_1\in S$,
\begin{equation}
\label{eq:greenrep}
\green(s_1)\cdot \E_{s_2\sim \rho}
f(s_2)\psi(s_2)\approx \transp{\1_{s_1}}\,\Delta^{-1} \psi
\end{equation}

If the state space is finite, an exact
solution is indeed $E=\R^{\#S}$ with a one-hot representation for $f$,
and the true matrix $\Delta^{-1}$ for $\green$, namely,
\begin{equation}
f(s_2)=\1_{s_2}/\rho(s_2),\qquad
(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}
\end{equation}
or equivalently, $f(s_2)=\1_{s_2}$ and
$(\green(s_1))_{s_2}=(\Delta^{-1})_{s_1s_2}/\rho(s_2)$. \footnote{The solution to
such representation problems is never unique because $E$ is an arbitrary
space and one may always make a change of basis in $E$, e.g., dividing
the $s_2$-coordinate by $\rho(s_2)$.}
But with infinite or continuous state spaces, we have to find a more
compact representation.

The reason for this form is the following: $E$ is supposed to represent
distributions or functions over states. $f$ is supposed to map $s_2$ to
something like
$\1_{s_2}$ (up to a factor $1/\rho$ to compensate for sampling
probabilities). 
In
particular, $\E_{s_2\sim \rho}f(s_2)\psi(s_2)$ is supposed to represent
the vector with components $\psi(s_2)$, namely, the function $\psi$. $\green$ is supposed to map each
$s_1$ to the
the occupation
distribution of the random walk starting from $s_1$, namely,
$\transp{\1_{s_1}}\sum_t P^t$, which is $\transp{\1_{s_1}}\,\Delta^{-1}$
(as usual I
omit constants\footnote{To deal with constants, we probably have to be
able to predict \emph{differences} $\psi(s_2)-\psi(s_2')$, or just use
relative value functions and remove $\psi(s_0)$ for some $s_0$.}). Then the scalar product in
$E$ represents integration/summation over the state space.

Of course we cannot directly learn $\green$ from \eqref{eq:greenrep},
since $\Delta^{-1}$ in the right-hand-side is exactly what we want to
learn. But $\Delta^{-1}$ is the unique solution to the Bellman equations \eqref{eq:greenbell}; after
applying \eqref{eq:greenbell} to an arbitrary state $s_1$ and test
function $\psi$, and
substituting the model $\green$ for $\Delta^{-1}$ via
$\eqref{eq:greenrep}$, the left Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,P\Delta^{-1}\psi
\qquad \Longrightarrow
\\
\green(s_1)\cdot\E_{s_2\sim \rho}
f(s_2)\psi(s_2)
\approx \psi(s_1)+ \E_{s'\sim P(s'|s_1)}\green(s')\cdot
\E_{s_2\sim \rho}\,
f(s_2)\psi(s_2)
\end{gather}
because $\transp{\1_{s_1}}\,P=\E_{s'\sim P(s'|s_1)}\transp{\1_{s'}}\,$.
Meanwhile,
the right Bellman equation becomes
\begin{gather}
\transp{\1_{s_1}}\,\Delta^{-1}\psi=\transp{\1_{s_1}}\,\psi+\transp{\1_{s_1}}\,\Delta^{-1}P\psi
\qquad \Longrightarrow
\\
\green(s_1)\cdot\E_{s_2\sim \rho}
f(s_2)\psi(s_2)
\approx \psi(s_1)+ \green(s_1)\cdot\E_{s_2\sim \rho}\E_{s'_2\sim P(s'_2|s_2)}
f(s_2)\psi(s'_2)
\end{gather}
(indeed, 
$P\psi$ is
represented by $\E_{s_2\sim \rho} f(s_2)(P\psi)(s_2)$, and
$(P\psi)(s_2)=\E_{s'_2\sim P(s'_2|s_2)}\psi(s'_2)$).

\emph{Both} these equations are samplable. Therefore, one may
parameterize $f$ and $\green$ by neural networks and train the
parameters. This works off-policy: we have not used that $\rho$ is the
invariant distribution, just some distribution on states we can sample
from.

Of course, we should apply this to relevant test functions $\psi$,
such as the reward function or any useful (predictive) features, namely,
any function we're interested in predicting the future values of. 
If just using a
random function $\psi$, there is very little signal to train
from. \footnote{An extreme case is to take a purely random $\psi$ at each
training step; then $\psi(s_1)$ and $\psi(s_2)$ are independent unless $s_1=s_2$,
which is a very unlikely event. This amounts to training with white noise
features. Formally it works thanks to the rare case $s_1=s_2$, but
variance is just terrible.}

Thus we may learn $\green$ via TD-like algorithms. We may even have
double sampling that works, thanks to featurization (using test functions
$\psi$), at least for the right equation. TODO should we featurize on the
left as well?

The value function is obtained with $\psi=R$, namely,
\begin{equation}
V(s_1)=\transp{\1_{s_1}}\Delta^{-1}R\approx
\green(s_1)\cdot\E_{s_2\sim \rho}
f(s_2)R(s_2)
\end{equation}
Of course it is not convenient to have to sample from $s_2$. Instead one
might want to just learn the representation $f(R)$, defined by abuse of
notation as
\begin{equation}
f(R)\deq \E_{s_2\sim \rho} \,f(s_2)R(s_2)
\end{equation}
(which can be computed online)
and then
\begin{equation}
V(s_1)\approx \green(s_1)\cdot f(R)
\end{equation}

More generally, we can learn the representation of any function $\psi$.
This could be useful if we want to train a few policy parameters to lead in
certain directions/lead to states with certain features.

Another equation that might be enforced is commutation between
$P$ and $\Delta^{-1}$; it is expressed similarly to the above, but I do
not know if there's a gain. It might transfer left to right Bellman
equations.

$\green$ is basically a model of the world (summed over time). The
difference with ordinary model-based is: $\green$ is trained as a fixed
point, considering all timescales at once. In ordinary model-based, the
model $\tilde P$ obviously commutes with its iterates ($\tilde P^t\tilde
P=\tilde P\tilde P^t$), but does not provide an easy and accurate
representation of the occupation measure $\tilde \Delta^{-1}=\sum_t \tilde P^t$. So even in
the model-based case, the above might be useful.

>>>>

\newcommand{\Dir}{\mathcal{D}}

@@

$G$ cannot be stored but one can store an approximate model. For instance
one could approximate $G$ via a neural network with input $s$ and output
in the policy parameter space. \footnote{The policy parameter space may
be quite large. This means the neural network should have a small
next-to-last layer. Indeed, with a next-to-last layer of size $r$,
the weights of the last layer are represented by $r$ policy parameters.
This corresponds to a rank-$r$ approximation of $G$,
as a sum of parameters times functions:
\begin{equation}
G\approx \sum_i \psi_i \phi_i(s)
\end{equation}
where each $\psi_i$ is a policy parameter, and each $\phi_i$ is
a function of the state, corresponding to a feature in the network.} Then the reverse
Bellman equation above should allows us to learn this model via
a TD-like algorithm, \emph{on-policy}. TODO: HOW? not clear how to
sample from $\phi P$ if $\phi$ is a function. Would need a model of
reverse transitions? \footnote{I think one has to learn $G\mu^{-1}$ and
the equation $G=\Id+GP$ rewrites as $\mu
(G\mu^{-1})\mu=\mu+\mu(G\mu^{-1})\mu P$ where everything is samplable.
However I fear an implicit double sampling in there; probably
featurize/dualize
by asking that this is satisfied for any test functions on the left and right.

Another way to exploit the equation on the right is to impose that $G$
and $P$ commute (indeed $G=\sum P^t$). This commutation equation should
transfer left-learning to right-learning. It can be written in samplable
terms
via $\mu(G\mu^{-1})(\mu P\psi)=(\mu P)(G \mu^{-1})(\mu
\psi)$ for any feature $\psi$. This apparently avoids double sampling!
}

@@
